Hybrid Threads for the Parrot Virtual Machine

### 2012-03-29 Paper.tex
Update to match the current version of the code + spelling fixes

The Perl 6 design process began with asking the Perl users for what they were expecting from the new version of the language. The very first feature that got asked for was well integrated multithreading support\cite{RFC1}. Perl 5 had two different implementations of thread support. In the first model, called 5005threads, data was shared by default and shared access to data had to be explicitely synchronized. This was similiar to the models used by languages such as C or Java. The implementation however suffered from data corruption and crashes and thus was not recommended for production use\cite{ThreadManual}. Perl 5.6 introduced the newer model called ithreads, mostly as a way to emulate fork() on Win32 platforms. Perl 5.8 exposed this API (Application Programming Interface) to the user of the programming language. In this new model, all data is copied to each thread and afterwards thread local. Data must be explicitely shared between threads. In other words, in Perl threads are not lightweight at all. They have severe impact on memory usage, writes to shared data are expensive and still not all features of the language are usable in threaded programs.

Being born at a time when Perl 6 still looked much more similiar to Perl 5 than it does nowadays, Parrot's threading support initially was very close to Perl's ithreads model. Previous attempts to change this into the more conventional model of data shared by default or implementing new technologies like Software Transactional Memory failed. For example Parrot has never supported running multiple threads and having garbage collection at the same time.

@@

Low level programming languages like C provide only the bare necessities, leaving the responsibility for preventing data corruption entirely to the user. A high level language like Perl 6 on the other hand offers complex and compund data types, handles garbage collection and a very dynamic object system. Even seeminlgy simple things like a method call can become very complex sequences. In a statically typed programming language the definition of a class is immutable. Thus calling a method on an object contains just the steps of determining the object's class, fetching the required method from this class and calling it. Calling the same method again can then even omit the first two steps since their results cannot change.

@@

The CPython implementation of the Python runtime uses a Global Interpreter Lock (GIL) to protect its internal consistency\cite{PythonThreadingManual}. This is a single lock taken whenever the interpreter executes Python bytecode. Because of this lock, only one thread can execute bytecode at any time so all built-in types and the object model are implicitely type safe. The drawback is that Python code cannot benefit from having multiple CPU cores available. However I/O operations and calls to external libraries are done without holding the GIL, so in applications with multiple I/O bound threads, there may still be a performance benefit from using multithreading.

To actually run Python code in parallel, multiple processes have to be used. The \textit{multiprocessing} module provides support for spawning processes using an API similar to the threading module\cite{PythonMultiProcessingManual}. Since processes may not dirctly access other processes' memory, the \textit{multiprocessing} module provides several means of communication between processes: \textit{Queues}, \textit{Pipes} and shared memory support.

@@

Parrot consists of the virtual machine (VM, also called interpreter), and various tools to facilitate the implementation of programming languages on top of the Parrot VM (the Parrot Compiler Toolkit). This paper concentrates on the virtual machine itself. The interpreter itself is written in C. Example code and testcases are written in PIR (Parrot Intermediate Representation) a high level assembly language abstracting away register allocations and function calling conventions.

Contrary to other widely used virtual machines like the JVM or the CLR which are stack based, Parrot mirrors contemporary hardware CPUs more closely by being register based. A stack based virtual machine usually fetches the operands of an operation from the top of a stack and puts the result back on top. Thus the operands chosen implicitely by ordering of the operations allowing the opcodes to be quite small. In a register based VM on the other hand each operation has to specify the operands explicitely. Compilers for stack machines are simpler because they do not have to care about register allocations and code is independent of prior or subsequent code\cite{VMShowdown}. The rationale behind giving up the simplicity of a stack based implementation is the hope of simplifying just in time compilation and improved performance of nested function and method calls.

@@

PMCs (PolyMorphic Container) are containers for all higher level types such as objects, arrays, hash tables or code. Thus they are very similar to Python's PythonObject types. PMCs are defined by C structs and are fully garbage collected. Their definition looks like:
\begin{CCode}
struct PMC {
    Parrot_UInt    flags;
    VTABLE         *vtable;             /* Pointer to vtable. */
    DPOINTER       *data;               /* Pointer to attribute structure. */
    PMC            *_metadata;          /* Pointer to metadata PMC. */
};
\end{CCode}
In short it contains a pointer to a type specific data structure and some meta data. The vtable is where the type's behaviour is defined. It contains a long list of function pointers forming Parrot's unified type interface. This interface is a union of numeric, string, array, hash and object like behaviours. For example the  \textit{get\_integer} function returns an integer value for the data type. For a simple \textit{int} it is the contained value. For an array it may be the number of elements. The \textit{find\_method} function makes user defined methods of objects possible.

In this way, a language implementor can define the basic types of the language and available operations on them. For each vtable entry, there is a corresponding op code in Parrot's bytecode. Thus an \textit{inc \$P0} instruction is calling the \textit{increment} vtable function of the PMC contained in the \textit{\$P0} register.

PMCs are implemented in .pmc files with method bodies written in C. These files get preprocessed to plain C before compliation.

@@

Control flow is modeled in Parrot using continuation passing. A continuation is a data structure containing the a state of a program at a given point in its execution. It contains all information needed to continue a program in a certain state, e.g. a call stack, the instruction pointer and contents of local variables. ``Calling'' a continuation means restoring the state encapsulated in the continuation.

@@

"Green threads" or "lightweight threads" are threads managed by the virtual machine instead of by the operating system. These threads only run pseudo parallel. The virtual machine contains a scheduler and support for preempting running tasks.
Preemption means to suspend a task regardless of what it is currently doing and probably resuming it later on.
This is an implementation of the many-to-one threading model very similiar to what an operating system running on a system with a single CPU core does.
%
Advantages of green threads are:
%
\begin{itemize}
\item They allow pseudo concurrent processing without endangering the interpreter's internal consistency.
\item Green threads are very light weight with low memory overhead and close to zero creation time.
\item They do not depend on OS threading support
\item The interpreter controls the point of preemption of a green thread.
\item The interpreter controls the exact scheduling policy and may allow the user to influence it or even take over completely.
\item Garbage collection can be implemented like in a single threaded process.
\item Critical sections can simply be protected by disabling the scheduler until the code is clear of the section.
\end{itemize}
%
Disadvantages include:
%
\begin{itemize}
\item They do not allow more than one CPU core to be used for computations.
\item Blocking calls like I/O block the whole interpreter including other green threads.
\item The interpreter has to contain logic and timers to control green threads.
\end{itemize}

@@

Following is a simple example to explain why concurrent processing can endanger the interpreter's internal consistency. The basic problem is concurrent writes to shared data. Assuming an implementation of an array class consisting of the field holding the data and the number of contained elements in a seperate member variable:
\begin{CCode}
pmclass ResizableIntegerArray auto_attrs provides array {
    ATTR INTVAL   size;      /* number of INTVALs stored in this array */
    ATTR INTVAL * int_array; /* INTVALs are stored here */
\end{CCode}
to append a new value, the array would have to read the current size, write the new value at the position size + 1 and then write the incremented size back into the member variable.
Now if two threads simultainously try to do this, it may happen that both read the same size, write to the same position (with one overwriting the value of the other) and write back the same incremented size. In this case one of the appended values would be lost.

In a more complicated example, the array could have to resize its data buffer to accomodate the new value. It would read the current size, allocate a new buffer, copy the values to the new buffer and then destroying the old one. Again if two threads try to do this simultainously, one of them could still be copying data, while the other is already at the point where it destroys the old buffer. This would lead to the copying thread accessing freed memory.

@@

This chapter describes how green threads were used to solve the following problems occuring when trying to implement threading support:
%
\begin{itemize}
\item How to ensure internal interpreter consistency when doing writes to shared data.
\item How to implement critical sections.
\item How to handle garbage collection.
\end{itemize}

In keeping the analogy of the interpreter being a software CPU, multithreading is implemented by having a separate interpreter with its own register set for each thread. When a user starts a new task, the scheduler first looks for an idle thread. If one can be found, the task is scheduled on the thread's interpreter. If none can be found, a new thread with a new interpreter is started. Parrot tries to optimize the number of utilized threads by creating at most one for each CPU core in the system. If more tasks are started than the maximum number of threads, the tasks are distributed evenly among the running interpreters. This is effectively an implementation of the N:M threading model.

@@

\textit{Parrot\_cx\_init\_scheduler} is called from \textit{Parrot\_interp\_initialize\_interpreter} whenever a new interpreter is created, e.g. on interpreter startup and whenever a new thread is created. It creates the scheduler PMC and sets up alarm singal handling.

@@

\textit{Parrot\_cx\_next\_task} contains the code to take the next task from the run queue and executing it. Before calling the task, it checks if there are other tasks still in the queue. Only if other tasks are waiting, task preemption gets enabled. Otherwise the current task would run until it finishes by itself, schedules other tasks or an alarm expires. this is an optimization reducing the runtime overhead of green threads in the important single tasking case to 0.

@@

When an exception is thrown, Parrot searches for a previously set up exception handler and executes it. As part of exception handling finalization, it cleans up, freeing any information no longer needed. The exception handler does not necessarily have to have been set up in the same runloop from which the exception has been thrown. Parrot uses the runloop id to identifiy such situations and free the nested runloops up to the level of the exception handler.

@@

\textit{Parrot\_thread\_outer\_runloop} is used as the thread's main function. It's very similiar to \textit{Parrot\_cx\_outer\_runloop}. The most important difference is the way it waits for pending alarms in case no tasks are to be run. While \textit{Parrot\_cx\_outer\_runloop} waits for an alarm signal to arrive, \textit{Parrot\_thread\_outer\_runloop} waits for a dummy to be written to its \textit{notifierfd} pipe by \textit{Parrot\_cx\_run\_scheduler}. In addition it is holding the \textit{interp->thread\_data->interp\_lock} lock while executing a task. This is the lock which \textit{Parrot\_cx\_schedule\_task} is taking during its call to \textit{Parrot\_thread\_schedule\_task} because the latter needs to create new objects local to the thread. As this could cause garbage collection to be triggered, the thread itself should not be executing code at the same time.

@@

As described in section~\ref{sec:PMCs} PMC source files to get preprocessed before compilation by a C compiler. This preprocessing gets done by the \textit{tools/build/pmc2c.pl} Perl script. This script got extended by implementing the \textit{Parrot::Pmc2c::PMC::Proxy} module which creates default implementations for all vtable functions not otherwise defined in the \textit{proxy.pmc} file. The default implementation for all writing functions would just call \textit{cant\_do\_write\_method} which creates a runtime exception. All other methods call the vtable method of the same name on the \textit{target} passing \textit{interp} as the current interpreter and all other parameters unchanged. If the method returns a PMC, another proxy object is created and wrapped around it so it can be safely returned to the caller.

@@

@@

Since each interpreter has its own garbage collector and consequenty its own memory areas and can only access its own data directly, the garbage collector can simply act as if there were no other threads. Thus the main work has been to clean up the interpreter structure itself and to replace references to global data by proxies. Only a few special cases had to be built into the GC itself. The global variable PMCNULL is used in many places in the code, but should be handled only by the main thread's garbage collector.

@@

Code segments and vtables are only marked by the main thread's garbage collector.

>>>>

The Perl 6 design process began with asking the Perl users for what they were expecting from the new version of the language. The very first feature that got asked for was well integrated multithreading support\cite{RFC1}. Perl 5 had two different implementations of thread support. In the first model, called 5005threads, data was shared by default and shared access to data had to be explicitly synchronized. This was similar to the models used by languages such as C or Java. The implementation however suffered from data corruption and crashes and thus was not recommended for production use\cite{ThreadManual}. Perl 5.6 introduced the newer model called ithreads, mostly as a way to emulate fork() on Win32 platforms. Perl 5.8 exposed this API (Application Programming Interface) to the user of the programming language. In this new model, all data is copied to each thread and afterwards thread local. Data must be explicitly shared between threads. In other words, in Perl threads are not lightweight at all. They have severe impact on memory usage, writes to shared data are expensive and still not all features of the language are usable in threaded programs.

Being born at a time when Perl 6 still looked much more similar to Perl 5 than it does nowadays, Parrot's threading support initially was very close to Perl's ithreads model. Previous attempts to change this into the more conventional model of data shared by default or implementing new technologies like Software Transactional Memory failed. For example Parrot has never supported running multiple threads and having garbage collection at the same time.

@@

Low level programming languages like C provide only the bare necessities, leaving the responsibility for preventing data corruption entirely to the user. A high level language like Perl 6 on the other hand offers complex and compound data types, handles garbage collection and a very dynamic object system. Even seemingly simple things like a method call can become very complex sequences. In a statically typed programming language the definition of a class is immutable. Thus calling a method on an object contains just the steps of determining the object's class, fetching the required method from this class and calling it. Calling the same method again can then even omit the first two steps since their results cannot change.

@@

The CPython implementation of the Python runtime uses a Global Interpreter Lock (GIL) to protect its internal consistency\cite{PythonThreadingManual}. This is a single lock taken whenever the interpreter executes Python bytecode. Because of this lock, only one thread can execute bytecode at any time so all built-in types and the object model are implicitly type safe. The drawback is that Python code cannot benefit from having multiple CPU cores available. However I/O operations and calls to external libraries are done without holding the GIL, so in applications with multiple I/O bound threads, there may still be a performance benefit from using multithreading.

To actually run Python code in parallel, multiple processes have to be used. The \textit{multiprocessing} module provides support for spawning processes using an API similar to the threading module\cite{PythonMultiProcessingManual}. Since processes may not directly access other processes' memory, the \textit{multiprocessing} module provides several means of communication between processes: \textit{Queues}, \textit{Pipes} and shared memory support.

@@

Parrot consists of the virtual machine (VM, also called interpreter), and various tools to facilitate the implementation of programming languages on top of the Parrot VM (the Parrot Compiler Toolkit). This paper concentrates on the virtual machine itself. The interpreter itself is written in C. Example code and test cases are written in PIR (Parrot Intermediate Representation) a high level assembly language abstracting away register allocations and function calling conventions.

Contrary to other widely used virtual machines like the JVM or the CLR which are stack based, Parrot mirrors contemporary hardware CPUs more closely by being register based. A stack based virtual machine usually fetches the operands of an operation from the top of a stack and puts the result back on top. Thus the operands chosen implicitly by ordering of the operations allowing the opcodes to be quite small. In a register based VM on the other hand each operation has to specify the operands explicitly. Compilers for stack machines are simpler because they do not have to care about register allocations and code is independent of prior or subsequent code\cite{VMShowdown}. The rationale behind giving up the simplicity of a stack based implementation is the hope of simplifying just in time compilation and improved performance of nested function and method calls.

@@

PMCs (PolyMorphic Container) are containers for all higher level types such as objects, arrays, hash tables or code. Thus they are very similar to Python's PythonObject types. PMCs are defined by C structs and are fully garbage collected. Their definition looks like:
\begin{CCode}
struct PMC {
    Parrot_UInt    flags;
    VTABLE         *vtable;             /* Pointer to vtable. */
    DPOINTER       *data;               /* Pointer to attribute structure. */
    PMC            *_metadata;          /* Pointer to metadata PMC. */
};
\end{CCode}
In short it contains a pointer to a type specific data structure and some meta data. The vtable is where the type's behavior is defined. It contains a long list of function pointers forming Parrot's unified type interface. This interface is a union of numeric, string, array, hash and object like behaviors. For example the  \textit{get\_integer} function returns an integer value for the data type. For a simple \textit{int} it is the contained value. For an array it may be the number of elements. The \textit{find\_method} function makes user defined methods of objects possible.

In this way, a language implementer can define the basic types of the language and available operations on them. For each vtable entry, there is a corresponding op code in Parrot's bytecode. Thus an \textit{inc \$P0} instruction is calling the \textit{increment} vtable function of the PMC contained in the \textit{\$P0} register.

PMCs are implemented in .pmc files with method bodies written in C. These files get preprocessed to plain C before compilation.

@@

Control flow is modeled in Parrot using continuation passing. A continuation is a data structure containing the state of a program at a given point in its execution. It contains all information needed to continue a program in a certain state, e.g. a call stack, the instruction pointer and contents of local variables. ``Calling'' a continuation means restoring the state encapsulated in the continuation.

@@

"Green threads" or "lightweight threads" are threads managed by the virtual machine instead of by the operating system. For better distinction green threads are henceforth referred to as "tasks". These tasks only run pseudo parallel. The virtual machine contains a scheduler and support for preempting running tasks.
Preemption means to suspend a task regardless of what it is currently doing and probably resuming it later on.
This is an implementation of the many-to-one threading model very similar to what an operating system running on a system with a single CPU core does.
%
Advantages of green threads are:
%
\begin{itemize}
\item They allow pseudo concurrent processing without endangering the interpreter's internal consistency.
\item Green threads are very light weight with low memory overhead and close to zero creation time.
\item They do not depend on OS threading support
\item The interpreter controls the point of preemption of a green thread.
\item The interpreter controls the exact scheduling policy and may allow the user to influence it or even take over completely.
\item Garbage collection can be implemented like in a single threaded process.
\item Critical sections can simply be protected by disabling the scheduler until the code is clear of the section.
\end{itemize}
%
Disadvantages include:
%
\begin{itemize}
\item They do not allow more than one CPU core to be used for computations.
\item Blocking calls like I/O block the whole interpreter including other green threads.
\item The interpreter has to contain logic and timers to control green threads.
\end{itemize}

@@

Following is a simple example to explain why concurrent processing can endanger the interpreter's internal consistency. The basic problem is concurrent writes to shared data. Assuming an implementation of an array class consisting of the field holding the data and the number of contained elements in a separate member variable:
\begin{CCode}
pmclass ResizableIntegerArray auto_attrs provides array {
    ATTR INTVAL   size;      /* number of INTVALs stored in this array */
    ATTR INTVAL * int_array; /* INTVALs are stored here */
\end{CCode}
to append a new value, the array would have to read the current size, write the new value at the position size + 1 and then write the incremented size back into the member variable.
Now if two threads simultaneously try to do this, it may happen that both read the same size, write to the same position (with one overwriting the value of the other) and write back the same incremented size. In this case one of the appended values would be lost.

In a more complicated example, the array could have to resize its data buffer to accommodate the new value. It would read the current size, allocate a new buffer, copy the values to the new buffer and then destroying the old one. Again if two threads try to do this simultaneously, one of them could still be copying data, while the other is already at the point where it destroys the old buffer. This would lead to the copying thread accessing freed memory.

@@

This chapter describes how green threads were used to solve the following problems occurring when trying to implement threading support:
%
\begin{itemize}
\item How to ensure internal interpreter consistency when doing writes to shared data.
\item How to implement critical sections.
\item How to handle garbage collection.
\end{itemize}

In keeping the analogy of the interpreter being a software CPU, multithreading is implemented by having a separate interpreter with its own register set for each thread. Thus the words ``thread'' and ``interpreter'' are used interchangeably as there is a 1:1 relationship between them. When a user starts a new task, the scheduler first looks for an idle thread. If one can be found, the task is scheduled on the thread's interpreter. If none can be found, a new thread with a new interpreter is started. Parrot tries to optimize the number of utilized threads by creating at most one for each CPU core in the system. If more tasks are started than the maximum number of threads, the tasks are distributed evenly among the running interpreters. This is effectively an implementation of the N:M threading model.

@@

\textit{Parrot\_cx\_init\_scheduler} is called from \textit{Parrot\_interp\_initialize\_interpreter} whenever a new interpreter is created, e.g. on interpreter startup and whenever a new thread is created. It creates the scheduler PMC and sets up alarm signal handling.

@@

\textit{Parrot\_cx\_next\_task} contains the code to take the next task from the run queue and executing it. Before calling the task, it checks if there are other tasks still in the queue. Only if other tasks are waiting, task preemption gets enabled. Otherwise the current task would run until it finishes by itself, schedules other tasks or an alarm expires. This is an optimization reducing the runtime overhead of green threads in the important single tasking case to 0.

@@

When an exception is thrown, Parrot searches for a previously set up exception handler and executes it. As part of exception handling finalization, it cleans up, freeing any information no longer needed. The exception handler does not necessarily have to have been set up in the same runloop from which the exception has been thrown. Parrot uses the runloop id to identify such situations and free the nested runloops up to the level of the exception handler.

@@

\textit{Parrot\_thread\_outer\_runloop} is used as the thread's main function. It's very similar to \textit{Parrot\_cx\_outer\_runloop}. The most important difference is the way it waits for pending alarms in case no tasks are to be run. While \textit{Parrot\_cx\_outer\_runloop} waits for an alarm signal to arrive, \textit{Parrot\_thread\_outer\_runloop} waits for a dummy to be written to its \textit{notifierfd} pipe by \textit{Parrot\_cx\_run\_scheduler}.

@@

As described in section~\ref{sec:PMCs} PMC source files to get preprocessed before compilation by a C compiler. This preprocessing gets done by the \textit{tools/build/pmc2c.pl} Perl script. This script got extended by implementing the \textit{Parrot::Pmc2c::PMC::Proxy} module which creates default implementations for all vtable functions not otherwise defined in the \textit{proxy.pmc} file. The default implementation for all writing functions would just call \textit{cant\_do\_write\_method} which creates a runtime exception.

All other methods call the vtable method of the same name on the \textit{target} passing the current interpreter as \textit{interp} and all other parameters unchanged. This causes all access to globals by the proxied function to go through the thread's interp and new PMCs created during the call to be allocated from the thread's memory pool. If the method returns a PMC from the target's interp, another proxy object has to be created and wrapped around it so it can be safely returned to the caller.

To differ between PMCs originating from the target's \textit{interp} and those created on the thread's interp during the call the GC is told to set the new \textit{PObj\_is\_new\_FLAG} on newly created PMCs. The \textit{PARROT\_THR\_FLAG\_NEW\_PMC} flag on the interp itself is used to communicate this requirement to the GC.

The alternative of calling the proxied function using the target's \textit{interp} would lead to concurrency issues with the garbage collector. Any call may cause garbage collection to run. Since the GC also scans the C stack the target's GC would find the thread's PMCs there and mark them confusing the thread's GC. Disabling the garbage collector would require a lock since the target's GC may already have started collecting.

\subsubsection{Sub}

The \textit{Sub} PMC represents executable subroutines. A \textit{Sub} does not only contain the code to execute but also the context in which to execute the code such as visible globals and namespaces. If a proxy to such a \textit{Sub} were created and \textit{invoke} called on it, the code would access this context directly since it belongs to the same interp as the proxied \textit{Sub} itself. Thus an op like \textit{get\_global} would fetch a global from an unproxied namespace and an unproxied global would be put into the target register. Since this is happening while running \textit{invoke} on the original \textit{Sub}, \textit{Proxy} cannot intercept the call and create a \textit{Proxy} for the result.

This is the reason why \textit{Parrot\_thread\_create\_proxy} does not create a \textit{Proxy} for a \textit{Sub} but uses \textit{Parrot\_thread\_create\_local\_sub} to create a copy on the thread's \textit{interp} with proxies for all PMC attributes like \textit{namespace\_stash} and \textit{ctx}.

@@

\subsection{The \textit{wait} op}

Using \textit{Tasks} to write to write to shared data makes such actions inherently asynchronous. This is not always what is needed by the implemented algorithm. For example when the shared variable is a lock, processing should continue as soon as it's acquired. The \textit{wait} op is used to wait for a task's completion. The waiting task is added to the waited for task's \textit{waiters} list and preempted immediately. When a task finishes, the all tasks in the \textit{waiters} list are scheduled again for execution. Since for each task a local copy is created on the target thread, the running task not only checks its own \textit{waiters} list but also its partner's.

If a task on the main thread was waiting for a task on another thread to finish and no other tasks are in the scheduler's queue on the main thread, the main thread would exit if no alarms are pending. To prevent this unintended exit all tasks are added to the scheduler's \textit{foreign\_tasks} list when they are scheduled on other threads. To end the program with other threads still running an explicit \textit{exit} op has to be issued.

@@

Since each interpreter has its own garbage collector and consequently its own memory areas and can only access its own data directly, the garbage collector can simply act as if there were no other threads. Thus the main work has been to clean up the interpreter structure itself and to replace references to global data by proxies. Only a few special cases had to be built into the GC itself. The global variable PMCNULL is used in many places in the code, but should be handled only by the main thread's garbage collector.

@@

Code segments and vtables are only marked by the main thread's garbage collector. Code segments are managed in Parrot using \textit{PackFile} structures. Besides the code itself these structures contain tables of constants used in the code. These constants are expanded to PMCs by the bytecode loader when the code is loaded from disk or put there directly by the PIR compiler. Since \textit{PackFiles} are no PMCs, the Proxy PMC cannot be used to shield them from the GC. Accessing the contained constants would therefore lead to unproxied PMCs being accessed by different threads. Since PackFiles may only be loaded by the main thread anyway, the bytecode loader and the compiler were changed to flag all \textit{PackFile} constants as \textit{shared}. They still are used as is on all threads, but the GC knows to only handle \textit{shared} PMCs on the main thread.

Scheduling tasks on other interpreters may cause garbage collection due to the need of creating proxy objects. To protect the garbage collector from concurrency issues, garbage collection is disabled during creation of the new \textit{Task} object and the proxies on the target thread's \textit{interp}. But since the target thread may itself be allocating memory or collecting garbage, the GC has to be protected by a lock as well. The \textit{interp->thread\_data->interp\_lock} lock is used for this. It is checked by all allocation and freeing functions which are also the only places where a garbage collection run may be triggered.
